---
title: "TCGA Cancer Analysis Report"
subtitle: "`r params$project`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    df_print: kable
params:
  project: "TCGA-UCEC"
  results_dir: "results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(knitr)

results_dir <- params$results_dir

# Display p-values in scientific notation instead of rounding them to 0
fmt_p <- function(p) ifelse(is.na(p), NA, formatC(p, format = "e", digits = 2))
```

This report assembles the outputs of `analysis.R` for **`r params$project`**
into a single document: differential expression, pathway enrichment, immune
signature scoring, and survival analysis.

## Differential Expression

```{r de-summary, results = "asis"}
de_path <- file.path(results_dir, "differential_expression_results.csv")
if (file.exists(de_path)) {
  de <- read.csv(de_path)
  n_up <- sum(de$change == "Up"); n_down <- sum(de$change == "Down")
  n_wald <- sum(!is.na(de$wald_padj) & de$wald_padj < 0.05)
  cat(n_up + n_down, " genes (", n_up, " up, ", n_down, " down) have |log2FC| significantly above ",
      "log2(1.5). For comparison, the conventional LFC = 0 test calls ", n_wald, " genes significant.\n",
      sep = "")
}
```

```{r de-table}
if (file.exists(de_path)) {
  de %>%
    filter(!is.na(padj)) %>%
    arrange(padj) %>%
    head(20) %>%
    mutate(pvalue = fmt_p(pvalue), padj = fmt_p(padj)) %>%
    select(gene_name, gene_id, log2FoldChange, log2FoldChange_shrunk, pvalue, padj) %>%
    kable(digits = 3, caption = paste(
      "Top 20 genes by adjusted p-value. p-values test H0: |log2FC| <= log2(1.5) (DESeq2 lfcThreshold),",
      "so padj < 0.05 means the fold change significantly exceeds 1.5x. Positive log2FC = higher in tumor",
      "(or MKI67-high for the fallback split). Shrunken values use apeglm on the original counts, so for",
      "genes with outlier samples they can differ from the tested estimate (e.g. HSPA6)."
    ))
} else {
  cat("No differential expression results found. Run analysis.R first.")
}
```

```{r volcano-plot}
volcano_path <- file.path(results_dir, "volcano_plot.png")
if (file.exists(volcano_path)) {
  knitr::include_graphics(volcano_path)
}
```

## Matched-Pair Sensitivity Analysis

The primary differential expression analysis above uses an **unpaired** design
(every tumor and normal sample treated as an independent observation), since
it lets every available sample contribute rather than only the smaller subset
of patients with both a tumor and a matched normal. This section fits a **separate,
secondary** model (`~ patient + group`) on just the matched-pair subset, as a
check on whether the primary analysis's top genes hold up once patient
identity is properly accounted for.

```{r paired-sensitivity, results = "asis"}
paired_comp_path <- file.path(results_dir, "paired_sensitivity_comparison.csv")
if (file.exists(paired_comp_path)) {
  paired_comp <- read.csv(paired_comp_path)
  lfc_cor <- cor(paired_comp$primary_lfc, paired_comp$paired_lfc, method = "spearman")
  primary_sig <- paired_comp[!is.na(paired_comp$primary_padj) & paired_comp$primary_padj < 0.05, ]
  pct_still_sig <- if (nrow(primary_sig) > 0) {
    100 * mean(!is.na(primary_sig$paired_padj) & primary_sig$paired_padj < 0.05)
  } else NA

  cat(
    "Spearman correlation between primary and paired log2 fold changes: ", round(lfc_cor, 3), ". ",
    round(pct_still_sig, 1), "% of the primary analysis's significant genes remain significant ",
    "(padj < 0.05) in the paired model.\n", sep = ""
  )

  paired_comp %>%
    filter(!is.na(primary_padj) & primary_padj < 0.05) %>%
    arrange(primary_padj) %>%
    head(20) %>%
    mutate(primary_padj = fmt_p(primary_padj), paired_padj = fmt_p(paired_padj)) %>%
    kable(digits = 3, caption = "Primary analysis's top 20 significant genes: log2FC and padj in both models")
} else {
  cat("Matched-pair sensitivity analysis not run (too few patients had both a tumor and matched normal sample, or the comparison isn't a tumor-vs-normal one).")
}
```

## HSPA (Hsp70) Gene Family

```{r hspa-expression}
hspa_de_path <- file.path(results_dir, "hspa_family_de_results.csv")
if (file.exists(hspa_de_path)) {
  hspa_de <- read.csv(hspa_de_path)

  hspa_sig <- hspa_de %>% filter(change != "Not significant") %>% arrange(padj)

  if (nrow(hspa_sig) > 0) {
    hspa_sig %>%
      mutate(pvalue = fmt_p(pvalue), padj = fmt_p(padj)) %>%
      select(gene_name, log2FoldChange, log2FoldChange_shrunk, pvalue, padj, change) %>%
      kable(
        digits = 4,
        caption = paste0(nrow(hspa_sig), " of ", nrow(hspa_de),
                          " HSPA family genes have |log2FC| significantly > log2(1.5) (threshold-test padj < 0.05)")
      )
  } else {
    cat("No HSPA family genes reached significance in this cohort's DE results.")
  }
} else {
  cat("HSPA expression check not yet run. See analysis.R.")
}
```

<details>
<summary>All 14 HSPA family genes (click to expand)</summary>

```{r hspa-expression-full}
if (file.exists(hspa_de_path)) {
  hspa_de %>%
    arrange(padj) %>%
    mutate(pvalue = fmt_p(pvalue), padj = fmt_p(padj)) %>%
    select(gene_name, log2FoldChange, log2FoldChange_shrunk, pvalue, padj, change) %>%
    kable(digits = 4, caption = "All HSPA family genes found in this dataset")
}
```

</details>

```{r hspa-plot}
hspa_plot_path <- file.path(results_dir, "hspa_family_plot.png")
if (file.exists(hspa_plot_path)) {
  knitr::include_graphics(hspa_plot_path)
}
```

```{r hspa-mutations}
hspa_mut_path <- file.path(results_dir, "hspa_mutation_frequency.csv")
if (file.exists(hspa_mut_path)) {
  hspa_mut <- read.csv(hspa_mut_path)
  if (nrow(hspa_mut) > 0) {
    kable(hspa_mut, caption = paste(
      "HSPA family genes: mutation frequency in this cohort.",
      "pct_tumors_mutated = MutatedSamples / tumors_assessed x 100."
    ))
  } else {
    cat("No HSPA family gene mutations found in this cohort.")
  }
} else {
  cat("HSPA mutation check not yet run. See mutation_analysis.R.")
}
```

```{r hspa-oncoplot}
hspa_onco_path <- file.path(results_dir, "hspa_oncoplot.png")
if (file.exists(hspa_onco_path)) {
  knitr::include_graphics(hspa_onco_path)
}
```

### HSPA Survival Analysis

```{r hspa-survival-table}
hspa_surv_path <- file.path(results_dir, "hspa_survival_results.csv")
if (file.exists(hspa_surv_path)) {
  hspa_surv <- read.csv(hspa_surv_path)
  hspa_surv %>%
    select(gene_name, n, n_events, cox_hr_per_sd, cox_p, cox_padj, cox_zph_p,
           adj_cox_hr_per_sd, adj_cox_p, adj_cox_padj, adj_cox_zph_p, logrank_p, logrank_padj) %>%
    kable(digits = 3, caption = paste(
      "HSPA genes, primary tumors: Cox HR per SD of VST expression (unadjusted and stage-adjusted where",
      "stage was available) and median-split log-rank test. padj = BH across HSPA genes tested.",
      "cox_zph_p / adj_cox_zph_p = proportional-hazards check (cox.zph global p); p < 0.05 means the",
      "constant-hazard-ratio assumption is questionable for that model, and the HR should be read with that caveat."
    ))
} else {
  cat("HSPA survival analysis not yet run.")
}
```

### HSPA Expression by TCGA Molecular Subtype

```{r hspa-subtype-plot}
subtype_plot_path <- file.path(results_dir, "hspa_by_subtype_plot.png")
if (file.exists(subtype_plot_path)) {
  knitr::include_graphics(subtype_plot_path)
} else {
  cat("HSPA-by-subtype analysis not yet run, no significant HSPA genes, or no subtype data for this cohort.")
}
```

```{r subtype-coverage}
cov_path <- file.path(results_dir, "subtype_coverage.csv")
if (file.exists(cov_path)) {
  read.csv(cov_path) %>%
    kable(caption = paste(
      "Molecular subtype coverage (primary tumors). Subtypes come from the 2013 TCGA marker paper;",
      "'Not assigned' tumors lacked data for integrative clustering and are a mixed group, not a subtype."
    ))
}
```

```{r hspa-subtype-kruskal}
kw_path <- file.path(results_dir, "hspa_subtype_kruskal.csv")
if (file.exists(kw_path)) {
  read.csv(kw_path) %>%
    arrange(kruskal_p) %>%
    mutate(kruskal_p = fmt_p(kruskal_p), kruskal_padj = fmt_p(kruskal_padj)) %>%
    kable(caption = "Kruskal-Wallis test of expression across assigned subtypes ('Not assigned' excluded), BH-corrected across genes")
}
```


## Pathway Enrichment (GSEA, Hallmark Gene Sets)

```{r gsea-table}
gsea_path <- file.path(results_dir, "gsea_hallmark_results.csv")
if (file.exists(gsea_path)) {
  gsea <- read.csv(gsea_path)
  n_sig <- sum(gsea$padj < 0.05, na.rm = TRUE)

  if (n_sig > 0) {
    gsea %>%
      filter(padj < 0.05) %>%
      arrange(padj) %>%
      mutate(padj = fmt_p(padj)) %>%
      select(pathway, NES, padj, size) %>%
      kable(digits = 3, caption = paste(n_sig, "significantly enriched pathways (padj < 0.05)"))
  } else {
    cat(
      "No pathways reached significance at padj < 0.05 for this comparison. ",
      "Below are the top 10 pathways by absolute NES regardless of significance."
    )
    gsea %>%
      arrange(desc(abs(NES))) %>%
      head(10) %>%
      mutate(padj = fmt_p(padj)) %>%
      select(pathway, NES, padj, size) %>%
      kable(digits = 3, caption = "Top 10 by NES (not statistically significant)")
  }
}
```

```{r gsea-plot}
gsea_plot_path <- file.path(results_dir, "gsea_hallmark_plot.png")
if (file.exists(gsea_plot_path)) {
  knitr::include_graphics(gsea_plot_path)
} else {
  cat("No GSEA plot available (likely no pathways passed the significance threshold used for plotting).")
}
```

## Immune Signature Scoring

Mean marker-gene z-score (VST expression) for major immune populations and the
Ayers IFN-gamma signature, computed across primary tumors only.

```{r immune-heatmap}
immune_heatmap_path <- file.path(results_dir, "immune_signature_heatmap.png")
if (file.exists(immune_heatmap_path)) {
  knitr::include_graphics(immune_heatmap_path)
}
```

```{r immune-table}
immune_path <- file.path(results_dir, "immune_signature_scores.csv")
if (file.exists(immune_path)) {
  immune <- read.csv(immune_path)
  head(immune, 10) %>%
    kable(digits = 2, caption = "Immune signature scores, first 10 samples")
}
```

## Survival Analysis

Exploratory: Kaplan-Meier curves (median split) and Cox models (continuous
expression) for the top differentially expressed gene, in primary tumors with
one sample per patient. Being differentially expressed versus normal tissue does
not imply prognostic value, so treat this as hypothesis-generating.

```{r survival-table}
top_surv_path <- file.path(results_dir, "survival_top_gene_results.csv")
if (file.exists(top_surv_path)) {
  read.csv(top_surv_path) %>%
    select(gene_name, n, n_events, cox_hr_per_sd, cox_p, cox_zph_p, adj_cox_hr_per_sd, adj_cox_p, adj_cox_zph_p, logrank_p) %>%
    kable(digits = 3, caption = paste(
      "Top DE gene: survival statistics. cox_zph_p / adj_cox_zph_p = proportional-hazards check",
      "(cox.zph global p); p < 0.05 flags that the constant-hazard-ratio assumption is questionable."
    ))
}
```

```{r survival-plot}
survival_path <- file.path(results_dir, "survival_km_plot.png")
if (file.exists(survival_path)) {
  knitr::include_graphics(survival_path)
} else {
  cat("No survival plot available.")
}
```

## Immune-hot vs. Immune-cold Classification

Tumors are split into tertiles of the Ayers 6-gene IFN-gamma signature
(IFNG, STAT1, IDO1, CXCL9, CXCL10, HLA-DRA): top tertile = immune-hot, bottom =
immune-cold. The classes are relative to this cohort, not absolute thresholds.

```{r hot-cold-table}
if (file.exists(immune_path)) {
  read.csv(immune_path) %>%
    mutate(hot_cold = factor(hot_cold, levels = c("Immune-cold", "Intermediate", "Immune-hot"))) %>%
    count(hot_cold, name = "n_tumors") %>%
    kable(caption = "Tumors per immune class")
}
```

```{r hot-cold-plot}
hot_cold_path <- file.path(results_dir, "immune_hot_cold_plot.png")
if (file.exists(hot_cold_path)) {
  knitr::include_graphics(hot_cold_path)
}
```

## Mutation / Tumor Mutational Burden

```{r mutation-section}
tmb_path <- file.path(results_dir, "tmb_summary.csv")
if (file.exists(tmb_path)) {
  tmb <- read.csv(tmb_path)
  tmb %>%
    summarise(
      n_samples = n(),
      median_per_Mb = median(total_perMB),
      q90_per_Mb = quantile(total_perMB, 0.9),
      n_over_10_per_Mb = sum(total_perMB > 10)
    ) %>%
    kable(digits = 2, caption = "Tumor mutational burden summary (38 Mb capture size assumed)")
} else {
  cat("Mutation analysis not yet run. See mutation_analysis.R.")
}
```

```{r oncoplot}
oncoplot_path <- file.path(results_dir, "oncoplot.png")
if (file.exists(oncoplot_path)) {
  knitr::include_graphics(oncoplot_path)
}
```

```{r tmb-immune-correlation}
corr_path <- file.path(results_dir, "tmb_immune_correlation.png")
if (file.exists(corr_path)) {
  knitr::include_graphics(corr_path)
}
```

### TMB and Immune Context by Molecular Subtype

The pooled TMB-immune correlation mixes differences *between* subtypes (POLE and
MSI tumors carry most of the high-TMB tail) with any relationship *within* a
subtype. The table below tests the association separately in each assigned
subtype. 

```{r tmb-subtype-table}
sub_cor_path <- file.path(results_dir, "tmb_immune_subtype_correlation.csv")
if (file.exists(sub_cor_path)) {
  read.csv(sub_cor_path) %>%
    mutate(spearman_p = fmt_p(spearman_p), spearman_padj = fmt_p(spearman_padj)) %>%
    kable(digits = 2, caption = "TMB vs. T cell signature within each assigned subtype (Spearman; >= 10 tumors; BH across subtypes; 'Not assigned' excluded)")
} else {
  cat("Subtype-level TMB analysis not available (run mutation_analysis.R; requires subtype data).")
}
```

```{r tmb-subtype-plot}
sub_plot_path <- file.path(results_dir, "tmb_immune_by_subtype.png")
if (file.exists(sub_plot_path)) {
  knitr::include_graphics(sub_plot_path)
}
```

```{r class-by-subtype}
class_sub_path <- file.path(results_dir, "immune_class_by_subtype.csv")
if (file.exists(class_sub_path)) {
  read.csv(class_sub_path) %>%
    mutate(
      hot_cold = factor(hot_cold, levels = c("Immune-cold", "Intermediate", "Immune-hot")),
      cell = paste0(n, " (", round(100 * fraction), "%)")
    ) %>%
    arrange(hot_cold) %>%
    select(molecular_subtype, hot_cold, cell) %>%
    tidyr::pivot_wider(names_from = hot_cold, values_from = cell) %>%
    kable(caption = "Immune class (IFN-gamma tertiles) by molecular subtype: n (% of subtype)")
}
```

---

