---
title: "TCGA Cancer Analysis Report"
subtitle: "`r params$project`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    df_print: kable
params:
  project: "TCGA-UCEC"
  results_dir: "results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(knitr)

results_dir <- params$results_dir
```

This report assembles the outputs of `analysis.R` for **`r params$project`**
into a single document: differential expression, pathway enrichment, immune
signature scoring, and survival analysis.

## Differential Expression

```{r de-table}
de_path <- file.path(results_dir, "differential_expression_results.csv")
if (file.exists(de_path)) {
  de <- read.csv(de_path)
  de %>%
    filter(!is.na(padj)) %>%
    arrange(padj) %>%
    head(20) %>%
    select(gene_id, log2FoldChange, pvalue, padj) %>%
    kable(digits = 4, caption = "Top 20 genes by adjusted p-value")
} else {
  cat("No differential expression results found. Run analysis.R first.")
}
```

```{r volcano-plot}
volcano_path <- file.path(results_dir, "volcano_plot.png")
if (file.exists(volcano_path)) {
  knitr::include_graphics(volcano_path)
}
```

## HSPA (Hsp70) Gene Family

```{r hspa-expression}
hspa_de_path <- file.path(results_dir, "hspa_family_de_results.csv")
if (file.exists(hspa_de_path)) {
  hspa_de <- read.csv(hspa_de_path)

  hspa_sig <- hspa_de %>% filter(change != "Not significant") %>% arrange(padj)

  if (nrow(hspa_sig) > 0) {
    hspa_sig %>%
      select(gene_name, log2FoldChange, pvalue, padj, change) %>%
      kable(
        digits = 4,
        caption = paste0(nrow(hspa_sig), " of ", nrow(hspa_de),
                          " HSPA family genes reached significance (padj < 0.05, |log2FC| > log2(1.5))")
      )
  } else {
    cat("No HSPA family genes reached significance in this cohort's DE results.")
  }
} else {
  cat("HSPA expression check not yet run. See analysis.R.")
}
```

<details>
<summary>All 14 HSPA family genes (click to expand)</summary>

```{r hspa-expression-full}
if (file.exists(hspa_de_path)) {
  hspa_de %>%
    arrange(padj) %>%
    select(gene_name, log2FoldChange, pvalue, padj, change) %>%
    kable(digits = 4, caption = "All HSPA family genes found in this dataset")
}
```

</details>

```{r hspa-plot}
hspa_plot_path <- file.path(results_dir, "hspa_family_plot.png")
if (file.exists(hspa_plot_path)) {
  knitr::include_graphics(hspa_plot_path)
}
```

```{r hspa-mutations}
hspa_mut_path <- file.path(results_dir, "hspa_mutation_frequency.csv")
if (file.exists(hspa_mut_path)) {
  hspa_mut <- read.csv(hspa_mut_path)
  if (nrow(hspa_mut) > 0) {
    kable(hspa_mut, caption = "HSPA family genes: mutation frequency in this cohort")
  } else {
    cat("No HSPA family gene mutations found in this cohort.")
  }
} else {
  cat("HSPA mutation check not yet run. See mutation_analysis.R.")
}
```

```{r hspa-oncoplot}
hspa_onco_path <- file.path(results_dir, "hspa_oncoplot.png")
if (file.exists(hspa_onco_path)) {
  knitr::include_graphics(hspa_onco_path)
}
```

### HSPA Survival Analysis

```{r hspa-survival-table}
hspa_surv_path <- file.path(results_dir, "hspa_survival_results.csv")
if (file.exists(hspa_surv_path)) {
  hspa_surv <- read.csv(hspa_surv_path)
  kable(hspa_surv, digits = 4, caption = "HSPA genes: log-rank survival test (High vs. Low expression)")
} else {
  cat("HSPA survival analysis not yet run.")
}
```

### HSPA Expression by TCGA Molecular Subtype

```{r hspa-subtype-plot}
subtype_plot_path <- file.path(results_dir, "hspa_by_subtype_plot.png")
if (file.exists(subtype_plot_path)) {
  knitr::include_graphics(subtype_plot_path)
} else {
  cat("HSPA-by-subtype analysis not yet run or no subtype data available for this cohort.")
}
```


## Pathway Enrichment (GSEA, Hallmark Gene Sets)

```{r gsea-table}
gsea_path <- file.path(results_dir, "gsea_hallmark_results.csv")
if (file.exists(gsea_path)) {
  gsea <- read.csv(gsea_path)
  n_sig <- sum(gsea$padj < 0.05, na.rm = TRUE)

  if (n_sig > 0) {
    gsea %>%
      filter(padj < 0.05) %>%
      arrange(padj) %>%
      select(pathway, NES, padj, size) %>%
      kable(digits = 4, caption = paste(n_sig, "significantly enriched pathways (padj < 0.05)"))
  } else {
    cat(
      "No pathways reached significance at padj < 0.05 for this comparison. ",
      "Below are the top 10 pathways by absolute NES regardless of significance."
    )
    gsea %>%
      arrange(desc(abs(NES))) %>%
      head(10) %>%
      select(pathway, NES, padj, size) %>%
      kable(digits = 4, caption = "Top 10 by NES (not statistically significant)")
  }
}
```

```{r gsea-plot}
gsea_plot_path <- file.path(results_dir, "gsea_hallmark_plot.png")
if (file.exists(gsea_plot_path)) {
  knitr::include_graphics(gsea_plot_path)
} else {
  cat("No GSEA plot available (likely no pathways passed the significance threshold used for plotting).")
}
```

## Immune Signature Scoring

Mean marker-gene z-score across samples for major immune and stromal cell
populations. 

```{r immune-heatmap}
immune_heatmap_path <- file.path(results_dir, "immune_signature_heatmap.png")
if (file.exists(immune_heatmap_path)) {
  knitr::include_graphics(immune_heatmap_path)
}
```

```{r immune-table}
immune_path <- file.path(results_dir, "immune_signature_scores.csv")
if (file.exists(immune_path)) {
  immune <- read.csv(immune_path)
  head(immune, 10) %>%
    kable(digits = 2, caption = "Immune signature scores, first 10 samples")
}
```

## Survival Analysis

Kaplan-Meier survival stratified on the top differentially expressed gene.

```{r survival-plot}
survival_path <- file.path(results_dir, "survival_km_plot.png")
if (file.exists(survival_path)) {
  knitr::include_graphics(survival_path)
} else {
  cat("No survival plot available.")
}
```

## Immune-hot vs. Immune-cold Classification

Uses the Epithelial signature score against the T cell signature score, to classify
each sample's immune context.

```{r hot-cold-plot}
hot_cold_path <- file.path(results_dir, "immune_hot_cold_plot.png")
if (file.exists(hot_cold_path)) {
  knitr::include_graphics(hot_cold_path)
}
```

## Mutation / Tumor Mutational Burden

```{r mutation-section}
tmb_path <- file.path(results_dir, "tmb_summary.csv")
if (file.exists(tmb_path)) {
  tmb <- read.csv(tmb_path)
  head(tmb, 10) %>%
    kable(digits = 2, caption = "Tumor mutational burden, first 10 samples")
} else {
  cat("Mutation analysis not yet run. See mutation_analysis.R.")
}
```

```{r oncoplot}
oncoplot_path <- file.path(results_dir, "oncoplot.png")
if (file.exists(oncoplot_path)) {
  knitr::include_graphics(oncoplot_path)
}
```

```{r tmb-immune-correlation}
corr_path <- file.path(results_dir, "tmb_immune_correlation.png")
if (file.exists(corr_path)) {
  knitr::include_graphics(corr_path)
}
```

---

